Probabilistic Forecasting of the Disturbance Storm Time Index: An Autoregressive Gaussian Process approach

### 2017-03-16 agutmpl.tex
Penultimate changes

%Talk about Burton and friends
For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA), in this case one hour ahead prediction of $Dst$ because it is the simplest model towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One of the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where Q(t) relates to the particle injection from the plasma sheet into the inner magnetosphere.

@@

Let us assume that we want to model a process in which a scalar quantity $y$ is specified as $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, $d$ is the dimensionality of the input space, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is Gaussian distributed noise with variance $\sigma^2$.

@@

\begin{align}
  \mathbf{X}  = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_n \end{array} \right)_{n \times d} \label{eq:feat} \\
  \vspace{2\baselineskip}
  \mathbf{y}  = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{n \times 1} \label{eq:labels}
\end{align}

@@

\begin{enumerate}
      \item $\mathbf{I}$: The $n \times n$ identity matrix.
      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,n$ : Kernel matrix constructed from all couples obtained from the training data.
      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,n ; j \in 1,\cdots,m$ : Cross kernel matrix constructed from all couples between training and test data points.
      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,m$: Kernel matrix constructed from all couples obtained from the test data.
\end{enumerate}

@@

For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them. Some common kernel functions used in machine learning are listed in table \ref{table:kernel}.

@@

\begin{align*}
  Q(\theta) & = - log(p(\mathbf{y}|\mathbf{X}, K_\theta)) \\
            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I}| - \frac{n}{2}log(2\pi) \\
  \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{n \times n}
\end{align*}

@@

The objective function $Q(\theta)$ in the general case can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I}$ which has a time complexity $O(n^3)$ as noted above. In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$. Some of the techniques used for model selection in the context of GPR include.

@@

\item Maximum Likelihood: This technique as outlined in \citet{Rasmussen:2005:GPM:1162254} is a form of \emph{gradient descent}. It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$. Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data.

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}. An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity and IMF $B_z$.

@@

\begin{align*}
       \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
        & \ \ \ \ \  V(t-1), \cdots, V(t-p_{v}+1),\\
        & \ \ \ \ \  B_{z}(t-1), \cdots, B_{z}(t-p_{b}+1))
\end{align*}

@@

Apart from selecting the kernel parameters, one also needs to choose appropriate values for the auto-regressive orders $p$ in the case of GP-AR and $p, p_v, p_b$ in the case of GP-ARX. For this purpose we use a set of 24 storm events listed in table \ref{table:validationstorms} and for every assignment of values to the model order, we perform model selection with the routines in table \ref{table:modelselection} and record the performance on this validation set.

@@

In the case of GP-AR we let the model order $p$ vary from $5$ to $12$ while for GP-ARX we vary the total model order $p_t = p + p_v + p_b$ vary from $3$ to $12$ and for each $p_t$ evaluate every possible combination of $p$, $p_v$ and $p_b$ such that $p_t = p + p_v + p_b$ and $p, p_{v}, p_b > 0$.

@@

After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in table \ref{table:teststorms}, which is the same list of storm events as used in \citet{Ji2012}.

@@

Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX. The results are represented as boxplots, in which a rectangle is drawn to represent the second and third quartiles, with a vertical line inside to indicate the median value while outlying points are shown as dots. In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.

@@

After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of table \ref{table:teststorms}. The results of these model evaluations are summarized in table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.

@@

      \item \emph{Gaussian Process} AR and ARX models give encouraging benefits in OSA prediction. Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data. If one compares the training sets used by all the models, one can appreciate that the models presented here need relatively small training and validations sets: the training set contains 243 instances, while the validation set contains 782 instances.

>>>>

%Talk about Burton and friends
For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms. More specifically, we focus on the \emph{one step ahead} (OSA) (in this case one hour ahead) prediction of $Dst$ because it is the simplest model towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature. One of the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where $Q(t)$ relates to the particle injection from the plasma sheet into the inner magnetosphere.

@@

Let us assume that we want to model a process in which a scalar quantity $y$ is specified as $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, $d$ is the dimensionality of the input space, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is zero mean Gaussian noise with variance $\sigma^2$.

@@

\begin{align}
  \mathbf{X}  = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_N \end{array} \right)_{N \times d} \label{eq:feat} \\
  \vspace{2\baselineskip}
  \mathbf{y}  = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{N \times 1} \label{eq:labels}
\end{align}

@@

\begin{enumerate}
      \item $\mathbf{I}$: The $N \times N$ identity matrix.
      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,N$ : Kernel matrix constructed from all couples obtained from the training data.
      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,N ; j \in 1,\cdots,M$ : Cross kernel matrix constructed from all couples between training and test data points.
      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,M$: Kernel matrix constructed from all couples obtained from the test data.
\end{enumerate}

@@

For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them. Some common kernel functions used in machine learning are listed in Table \ref{table:kernel}.

@@

\begin{align*}
  Q(\theta) & = - log(p(\mathbf{y}|\mathbf{X}, K_\theta)) \\
            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}| - \frac{N}{2}log(2\pi) \\
  \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{N \times N}
\end{align*}

@@

The objective function $Q(\theta)$ in the general case can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above. In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$. Some of the techniques used for model selection in the context of GPR include.

@@

\item Maximum Likelihood: This technique as outlined in \citet{Rasmussen:2005:GPM:1162254} is a form of \emph{gradient descent}. It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$. Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}. An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.

@@

\begin{align*}
       \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
        & \ \ \ \ \  V_{sw}(t-1), \cdots, V_{sw}(t-p_{v}+1),\\
        & \ \ \ \ \  B_{z}(t-1), \cdots, B_{z}(t-p_{b}+1))
\end{align*}

@@

Apart from selecting the kernel parameters, one also needs to choose appropriate values for the auto-regressive orders $p$ in the case of GP-AR and $p, p_v, p_b$ in the case of GP-ARX. For this purpose we use a set of 24 storm events listed in Table \ref{table:validationstorms} and for every assignment of values to the model order, we perform model selection with the routines in Table \ref{table:modelselection} and record the performance on this validation set.

@@

In the case of GP-AR we let the model order $p$ vary from $5$ to $12$ while for GP-ARX we let the total model order $p_t = p + p_v + p_b$ vary from $3$ to $12$ and for each $p_t$ evaluate every possible combination of $p$, $p_v$ and $p_b$ such that $p_t = p + p_v + p_b$ and $p, p_{v}, p_b > 0$.

@@

After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in Table \ref{table:teststorms}, which is the same list of storm events as used in \citet{Ji2012}.

@@

Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX. The results are represented as boxplots, in which a rectangle is drawn to represent the first and third quartiles, with a vertical line inside to indicate the median value while outlying points are shown as dots. In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.

@@

After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of Table \ref{table:teststorms}. The results of these model evaluations are summarized in Table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.

@@

      \item \emph{Gaussian Process} AR and ARX models give encouraging benefits in OSA prediction. Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data. If one considers the size of the data used in our study, one can appreciate that the models presented here need relatively small training and validations sets: the training set contains 243 instances, while the validation set contains 782 instances.
